Secure Composition of Robust and Optimising Compilers

### 2023-09-21 csf/main.tex
fix cleverref
Remove larger chunks of text and fix some figures.

@@

Properties $\pi$ are sets of traces of admissible program behaviours, ascribing what said property considers valid.
The set of all properties can be partitioned into different {\em classes} ($\class$), i.e., safety, liveness, and neither safety nor liveness~\cite{clarkson2008hyper}.
A class is simply a set of properties and for the class of safety properties, it is decidable whether a trace satisfies a safety property with just a finite trace prefix.
As an example, consider a trace describing an interaction with a memory where the deallocation of an address $\loc$ precedes a read at that address in memory: $\ev{Dealloc\ \loc}\cdot\ev{Read\ \loc\ 1729}\cdot\dots$.
This program behaviour is insecure with respect to a canonical notion of (temporal) memory safety dictating no use-after-frees of pointers~\cite{nagarakatte2010cets,azevedo2018meaningsofms}, because it reads from a memory location that was freed already.
The previous finite trace prefix is enough to decide that the trace does not satisfy \gls*{tms} and there is no way to append events to this prefix which would result in the trace being admissible.
In the following, the execution of a whole program $\wholeprogvar$ that terminates in state $\runtimetermvar$ according to the language semantics and produces trace $\trace$ is written as $\progstepto{\wholeprogvar}{\runtimetermvar}{\trace}$.
With this, property satisfaction is defined as follows:
\bul{whole programs $\wholeprogvar$ satisfy a property $\pi$} iff \iul{$\wholeprogvar$ yields a trace $\trace$} such that \oul{$\trace$ satisifies $\pi$} (\Cref{def:propsat}).

\begin{definition}[Property Satisfaction]\label{def:propsat}
  \bul{$\sat{\progvar}{\pi}$}
  $\isdef$
  if \iul{$\progstepto{\progvar}{\runtimetermvar}{\trace}$},
  then \oul{$\trace\in\pi$}.
\end{definition}

@@

\begin{definition}[Robust Satisfaction]\label{def:proprsat}
  \bul{$\rsat{\progvar}{\pi}$}
  $\isdef$ \iul{$\forall \contextvar$, if $\link{\contextvar}{\progvar}=\wholeprogvar$}, then \oul{$\sat{\wholeprogvar}{\pi}$}.
\end{definition}

\begin{exampleenv}[Double Free in Bluetooth Subsystem]
  Consider \texttt{CVE-2021-3564}~\cite{doublefree-bluetooth}, one of many submissions for a double-free vulnerability.
  The vulnerability arises due to a race condition where the context-level function \texttt{hci\_cmd\_work} was not expected to behave maliciously, since it resides in the same source-code repository where the vulnerability occurs.
  Nevertheless, the component-level code of \texttt{hci\_dev\_do\_open} is linked with \texttt{hci\_cmd\_work} and does not atomically check whether a pointer has been freed already:
  Therefore, \texttt{hci\_dev\_do\_open} does not satisfy the no-double-frees property robustly, since there is an implementation for \texttt{hci\_cmd\_work} that leads to a violation of that property when linked with \texttt{hci\_dev\_do\_open}.
\end{exampleenv}

@@

\begin{definition}[\glsfirst*{tms}]\label{def:trace:tmsdef}
  \scriptsize$$
  \tmssafe:=\left\{\trace \left| \begin{array}{rcl}
                              \ev{Alloc\ \loc\ n;\sandboxtag;\securitytag}&\le_{\trace}&\ev{Dealloc\ \loc;\sandboxtag;\securitytag'} \\
                              \ev{Use\ \loc\ n;\sandboxtag;\securitytag}&\le_{\trace}&\ev{Dealloc\ \loc;\sandboxtag;\securitytag'} \\
                              \text{if }\ev{Alloc\ \loc\ n;\sandboxtag;\securitytag}\text{ in }\trace&\text{then}&\ev{Dealloc\ \loc;\sandboxtag;\securitytag'}\text{ in }\trace \\
                              \text{at most one }\ev{Dealloc\ \loc;\sandboxtag;\securitytag}&\text{in}&\trace \\
                              \text{at most one }\ev{Alloc\ \loc\ n;\sandboxtag;\securitytag}&\text{in}&\trace \\
                            \end{array}\right.\right\}
  $$
\end{definition}
Hereby, the notation $\event[_{1}]\le_{\trace}\event[_{2}]$ means that if $\event[_{1}]$ is in $\trace$ and if $\event[_{2}]$ is in $\trace$, then $\event[_{1}]$ appears before $\event[_{2}]$.

@@

Monitors enforce safety properties by accepting or rejecting traces, i.e., if they reject a trace, the trace does not satisfy the property they check.
Since reasoning on monitors is easier than directly on just traces, this section presents a monitor for each of the previously shown safety properties (\Cref{subsec:propdefs}).
%\MPpost{
%	this is fine, but it'd be nice to provide more insights on why it's easier
%}
To lessen the burden when proving that a monitor accepts the trace of a program execution, each monitor uses a custom trace model that contains only the relevant information related to the property the monitor checks.
To go from specification traces $\trace$ to monitor-level traces $\abstrace$, each property $\pi$ has an associated event agreement relation $\traceagree[]{\pi}{\event}{\absevent}$.
\Cref{fig:stuttereq} shows how the event agreement is lifted to traces.
\begin{figure}[!ht]
  \judgbox{\traceagree{\pi}{\trace}{\abstrace}}{,,Specification-level trace $\trace$ agrees with monitor-level\\trace $\abstrace$ with respect to property $\pi$.''}
  %
  \typerule{traceagree-empty}{}{
    \traceagree{\pi}{\hole{\cdot}}{\hole{\cdot}}
  }{traceagree-empty}
  %
  \typerule{traceagree-ign-L}{
    \traceagree{\pi}{\trace}{\abstrace}
  }{
    \traceagree{\pi}{\emptyevent\cdot\trace}{\abstrace}
  }{traceagree-ign-L}
  %
  \typerule{traceagree-ign-R}{
    \traceagree{\pi}{\trace}{\abstrace}
  }{
    \traceagree{\pi}{\trace}{\absev{\emptyevent}\cdot\abstrace}
  }{traceagree-ign-R}
  %
  \typerule{traceagree-cons}{
    \traceagree[]{\pi}{\event}{\absevent} &
    \traceagree{\pi}{\trace}{\abstrace}
  }{
    \traceagree{\pi}{\event\cdot\trace}{\absevent\cdot\abstrace}
  }{traceagree-cons}
  \caption{Trace-Agreement relation that equates specification-level traces with monitor-level traces.}\label{fig:stuttereq}
\end{figure}
The trace agreement is the same for all properties $\pi$ up to the use of the event agreement in \Cref{tr:traceagree-cons}.
With agreements, this section defines monitor satisfaction for traces and then it proves that monitor satisfaction implies property satisfaction.
To this end, monitor satisfaction is defined as follows.
\bul{A specification trace $\trace$ monitor-satisfies property $\pi$} iff there exists a (final) monitor state $\monitor$ and an abstract trace $\abstrace$ such that the \iul{specification trace $\trace$ agrees with abstract trace $\abstrace$} and \oul{the initial monitor\footnote{In this paper, for all monitors, the initial monitor state is denoted as $\emptyset$.} can step to the (final) monitor state $\monitor$ with abstract trace $\abstrace$}.

@@

\begin{exampleenv}[A program not satisfying \gls*{tms}]\label{ex:strncpy:cpp11}
  Consider the following \texttt{C++11} library that calls \texttt{strncpy} (\Cref{ex:strncpy}) and prints the result to the standard output stream.

  \begin{lstlisting}[language=c++,basicstyle=\small\ttfamily,mathescape,commentstyle=\color{gray}]
int hello() {
  char* greet = new char[11] { "Hello CSF!" };
  // greet address ${\color{gray}\loc_{x}}$
  char* to = new char[11];
  // to address ${\color{gray}\loc_{y}}$
  strncpy(11, to, greetings);
  delete to;
  printf("%cSF\n", to[6]);
}
  \end{lstlisting}
  Up to the body of \texttt{printf}, the program execution yields the specification trace $\ev{Alloc}\ \loc[_{x}]\ 12;\comp;\unlock\cdot\ev{Alloc}\ \loc[_{y}]\ 12;\comp;\unlock\cdot\ev{Use}\ \loc_{x}\ 0;\comp;\unlock\cdot\ev{Use}\ \loc_{x}\ 0;\comp;\unlock\cdot\ev{Use}\ \loc_{y}\ 0;\comp;\unlock\cdot\ev{Use}\ \loc_{x}\ 1;\comp;\unlock\cdot\ldots\cdot\ev{Use}\ \loc_{x}\ 12;\comp;\unlock\cdot\ev{Dealloc}\ \loc[_{y}];\comp;\unlock\cdot\ev{Use}\ \loc_{y}\ 6;\comp;\unlock$.
  Relating this trace to abstract monitor events yields $\abstrace=\absev{Alloc}\ \loc[_{x}]\ \comp\cdot\absev{Alloc}\ \loc[_{y}]\ \comp\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Use}\ \loc[_{y}]\ \comp\cdot\ldots\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Dealloc}\ \loc[_{y}]\ \comp\cdot\absev{Use}\ \loc[_{y}]\ \comp$.
  Remembering the definition of \texttt{strncpy} (\Cref{ex:strncpy}), observe that it does not deallocate its arguments.
  Even though the trace contains an out-of-bounds access right before returning from \texttt{strncpy}, this is no concern for \gls*{tms}, since the location $\loc[_{x}]$ is still allocated.
  However, having returned from \texttt{strncpy}, the \texttt{greet} function continues and deallocates $\loc[_{y}]$ whose subsequent use in the \texttt{printf} call is a use-after-free bug.

  The fix would be to \lstinline[language=c++,basicstyle=\small\ttfamily]|delete greetings| instead of \texttt{to} and add a \lstinline[language=c++,basicstyle=\small\ttfamily]|delete to| after the \texttt{printf} call, which leads to the abstract monitor trace $\abstrace[']=\absev{Alloc}\ \loc[_{x}]\ \comp\cdot\absev{Alloc}\ \loc[_{y}]\ \comp\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Use}\ \loc[_{y}]\ \comp\cdot\ldots\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Dealloc}\ \loc[_{x}]\ \comp\cdot\absev{Use}\ \loc[_{y}]\ \comp$.
  It follows that $\montmssafe{\abstrace[']}$ and from \Thmref{lem:mon:tmssafe}, it follows that the program satisfies \Thmref{def:trace:tmsdef}, even though the program still violates \gls*{sms}.
\end{exampleenv}

@@

\begin{exampleenv}[Normal invocation of \texttt{strncpy}]\label{ex:strncpy:sms}
  Consider the insecure \texttt{strncpy} function from \Cref{ex:strncpy} with a context \texttt{strncpy(2, x, y)}, where \texttt{x} and \texttt{y} are pointers to valid regions of memory with allocated space for exactly two cells and do not contain the null-terminating character \texttt{'\textbackslash 0'}.
  For the sake of this example, the pointers have been allocated by the component and passed to the context.
  The loop of \texttt{strncpy} will copy exactly two cells and then check the loop condition for the last time.
  At that stage, the induction variable \texttt{i} is equal to \texttt{2} and, unfortunately, the order of checks is such that first the cell \texttt{x[i]} is read prior to bounds checking \texttt{i < n}.
  Because of this, there is an out-of-bounds memory access right before exiting the function.
  This is also visible on the trace, which can be sketched as $\ldots\cdot\ev{Alloc}\ \loc_{x}\ 2;\comp;\unlock\cdot\ldots\cdot\ev{Alloc}\ \loc_{y}\ 2;\comp;\unlock\cdot\ldots\cdot\ev{Use}\ \loc_{x}\ 0;\comp;\unlock\cdot\ev{Use}\ \loc_{y}\ 0;\comp;\unlock\cdot\ev{Use}\ \loc_{x}\ 1;\comp;\unlock\cdot\ev{Use}\ \loc_{y}\ 1;\comp;\unlock\cdot\ev{Use}\ \loc_{x}\ 2;\comp;\unlock\cdot\ldots$, where $\loc_{x}$ and $\loc_{y}$ are the memory addresses associated to \texttt{x} and \texttt{y}, respectively.
  Omitting the events for all ,,$\ldots$'' for sake of brevity, the abstract monitor trace of this is $\absev{Alloc}\ \loc_{x}\ \comp\ 2\cdot\absev{Alloc}\ \loc_{y}\ \comp\ 2\cdot\absev{Use}\ \loc_{x}\ \comp\ 0\cdot\absev{Use}\ \loc_{y}\ \comp\ 0\cdot\absev{Use}\ \loc_{x}\ \comp\ 1\cdot\absev{Use}\ \loc_{y}\ \comp\ 1\cdot\absev{Use}\ \loc_{x}\ \comp\ 2$.

  After the allocation events, the state of the monitor is $\left\{(\loc_{x};\comp;2),(\loc_{y};\comp;2)\right\}$.
  All uses up to the last are accepted by the monitor, but the last event does not satisfy the premise $2<2$ in \Cref{tr:sms-use}.
  Therefore, the whole program (\texttt{strncpy} linked with this kind of context) is not \gls*{sms}.
\end{exampleenv}

@@

    \begin{tikzpicture}[scale=0.6, every node/.style={transform shape}]

@@

      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\small\ttfamily]

@@

      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\small\ttfamily]

@@

\subsection{Secure Upper and Lower Composition}\label{sec:other-compos}
Besides sequential composition, there are two other compositions, namely an {\em upper}, i.e., a compiler that takes multiple inputs and yields one output, and a {\em lower} composition, i.e., a compiler that takes one input and yields multiple outputs.
\bul{Define the upper composition $\cc{\src{L}+\obj{L}}{\trg{L}}$} as follows:
Given a program \texttt{p}, its compiled counterpart is obtained by \rul{plugging \texttt{p} into $\cc{\src{L}}{\trg{L}}$ if $\texttt{p}\in\src{L}$} or by \oul{plugging \texttt{p} into $\cc{\obj{L}}{\trg{L}}$ if $\texttt{p}\in\obj{L}$}.
\begin{definition}[Upper Composition]
  \bul{$\cc{\src{L}+\obj{L}}{\trg{L}}$} $\isdef$ 
  % Given \texttt{p}, yield
  $$\lambda \texttt{p}\ldotp
  \left\{\mbox{\begin{tabular}{c}
    {\rul{if $\texttt{p}\in\src{L}$, then $\cc{\src{L}}{\trg{L}}(\texttt{p})$}} \\
    \mbox{\oul{if $\texttt{p}\in\obj{L}$, then $\cc{\obj{L}}{\trg{L}}(\texttt{p})$}} \\
  \end{tabular}}\right.$$
%
%
%  $\begin{cases}
%                                                                            \rul{\cc{\src{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\src{L}} \\
%                                                                            \cc{\obj{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\obj{L}
%                                                                          \end{cases}$
\end{definition}

@@

Dually, the {\em lower} composition is concerned about compilers that accept the same source but yield different target languages. %and the same theoretical results apply.
\bul{Define the lower composition $\cc{\src{L}}{\trg{L}+\obj{L}}$} as follows:
Given a program $\src{p}$, its compiled counterpart is obtained by \rul{plugging $\src{p}$ into $\cc{\src{L}}{\trg{L}}$} or by \oul{plugging \texttt{p} into $\cc{\src{L}}{\obj{L}}$}, respectively, \pul{based on the internal decision}.
\begin{definition}[Lower Composition]
  \bul{$\cc{\src{L}}{\trg{L}+\obj{L}}$} $\isdef$ 
  % Given $L$ and $\src{p}$, yield
  $$\lambda \src{p}, L\ldotp
  \left\{\mbox{\begin{tabular}{c}
    {\pul{if $L=\trg{L}$, then} \rul{$\cc{\src{L}}{\trg{L}}(\src{p})$}} \\
    \mbox{\pul{if $L=\obj{L}$, then} \oul{$\cc{\src{L}}{\obj{L}}(\src{p})$}} \\
  \end{tabular}}\right.$$
%
%
%  $\begin{cases}
%                                                                            \rul{\cc{\src{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\src{L}} \\
%                                                                            \cc{\obj{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\obj{L}
%                                                                          \end{cases}$
\end{definition}

@@

% This paper considers \gls*{cct} as an architecture-dependent property\MPin{can we justify this}, so
Writing code attaining \gls*{cct} should not be of the programmer's concerns~\cite{cauligi2019fact}.
Such consideration is also backed up by architecture providing a data (operand) independent timing mode, such as processors by Arm~\cite[p.~543]{arm-refman} and Intel~\cite[p.~80]{intel-refman}.
In spirit of this, language $\Lscct$ allows violating \gls*{cct} by emitting events on, e.g., branching and division, that contain secrets (\Cref{subsec:lscct}), but provides a way to read and write to a {\em model--specific register} that enables a ``\gls*{cct}-mode''.

>>>>

% allow page breaks inside multiline alignment displays
\allowdisplaybreaks

\Crefname{exampleenv}{Example}{Examples}

@@

Properties $\pi$ are sets of traces of admissible program behaviours, ascribing what said property considers valid.
The set of all properties can be partitioned into different {\em classes} ($\class$), i.e., safety, liveness, and neither safety nor liveness~\cite{clarkson2008hyper}.
A class is simply a set of properties and for the class of safety properties, it is decidable whether a trace satisfies a safety property with just a finite trace prefix.
The compositionality framework (\Cref{sec:compcomp}) presented in this paper considers arbitrary classes, while the case-study (\Cref{sec:casestud:rtp}) fixes them to concrete instances.
As an example, consider a trace describing an interaction with a memory where the deallocation of an address $\loc$ precedes a read at that address in memory: $\ev{Dealloc\ \loc}\cdot\ev{Read\ \loc\ 1729}\cdot\dots$.
This program behaviour is insecure with respect to a canonical notion of (temporal) memory safety dictating no use-after-frees of pointers~\cite{nagarakatte2010cets,azevedo2018meaningsofms}, because it reads from a memory location that was freed already.
The previous finite trace prefix is enough to decide that the trace does not satisfy \gls*{tms} and there is no way to append events to this prefix which would result in the trace being admissible.
In the following, the execution of a whole program $\wholeprogvar$ that terminates in state $\runtimetermvar$ according to the language semantics and produces trace $\trace$ is written as $\progstepto{\wholeprogvar}{\runtimetermvar}{\trace}$.
With this, property satisfaction is defined as follows:
\bul{whole programs $\wholeprogvar$ satisfy a property $\pi$} iff \iul{$\wholeprogvar$ yields a trace $\trace$} such that \oul{$\trace$ satisifies $\pi$} (\Cref{def:propsat}).

\begin{definition}[Property Satisfaction]\label{def:propsat}
  \bul{$\sat{\progvar}{\pi}$}
  $\isdef$
  if \iul{$\forall\runtimetermvar\ \trace,\progstepto{\progvar}{\runtimetermvar}{\trace}$},
  then \oul{$\trace\in\pi$}.
\end{definition}

@@

\begin{definition}[Robust Satisfaction]\label{def:proprsat}
  \bul{$\rsat{\progvar}{\pi}$}
  $\isdef$ \iul{$\forall \contextvar\ \wholeprogvar$, if $\link{\contextvar}{\progvar}=\wholeprogvar$}, then \oul{$\sat{\wholeprogvar}{\pi}$}.
\end{definition}

% \begin{exampleenv}[Double Free in Bluetooth Subsystem]
%   Consider \texttt{CVE-2021-3564}~\cite{doublefree-bluetooth}, one of many submissions for a double-free vulnerability.
%   The vulnerability arises due to a race condition where the context-level function \texttt{hci\_cmd\_work} was not expected to behave maliciously, since it resides in the same source-code repository where the vulnerability occurs.
%   Nevertheless, the component-level code of \texttt{hci\_dev\_do\_open} is linked with \texttt{hci\_cmd\_work} and does not atomically check whether a pointer has been freed already:
%   Therefore, \texttt{hci\_dev\_do\_open} does not satisfy the no-double-frees property robustly, since there is an implementation for \texttt{hci\_cmd\_work} that leads to a violation of that property when linked with \texttt{hci\_dev\_do\_open}.
% \end{exampleenv}

@@

\begin{definition}[\glsfirst*{tms}]\label{def:trace:tmsdef}
  \footnotesize$$
  \tmssafe:=\left\{\trace \left| \begin{array}{rcl}
                              \ev{Alloc\ \loc\ n;\sandboxtag;\securitytag}&\le_{\trace}&\ev{Dealloc\ \loc;\sandboxtag;\securitytag'} \\
                              \ev{Use\ \loc\ n;\sandboxtag;\securitytag}&\le_{\trace}&\ev{Dealloc\ \loc;\sandboxtag;\securitytag'} \\
                              \text{if }\ev{Alloc\ \loc\ n;\sandboxtag;\securitytag}\text{ in }\trace&\text{then}&\ev{Dealloc\ \loc;\sandboxtag;\securitytag'}\text{ in }\trace \\
                              \text{at most one }\ev{Dealloc\ \loc;\sandboxtag;\securitytag}&\text{in}&\trace \\
                              \text{at most one }\ev{Alloc\ \loc\ n;\sandboxtag;\securitytag}&\text{in}&\trace \\
                            \end{array}\right.\right\}
  $$
\end{definition}
Hereby, the notation $\event[_{1}]\le_{\trace}\event[_{2}]$ means that if $\event[_{1}]$ is in $\trace$ and if $\event[_{2}]$ is in $\trace$, then $\event[_{1}]$ appears before $\event[_{2}]$.

@@

Monitors enforce safety properties by accepting or rejecting traces, i.e., if they reject a trace, the trace does not satisfy the property they check.
Since reasoning on monitors is easier than directly on just traces, this section presents a monitor for each of the previously shown safety properties (\Cref{subsec:propdefs}).
%\MPpost{
%	this is fine, but it'd be nice to provide more insights on why it's easier
%}
To lessen the burden when proving that a monitor accepts the trace of a program execution, each monitor uses a custom trace model that contains only the relevant information related to the property the monitor checks.
To go from specification traces $\trace$ to monitor-level traces $\abstrace$, each property $\pi$ has an associated event agreement relation $\traceagree[]{\pi}{\event}{\absevent}$.
Lifting the event agreement to traces works as expected, i.e., the empty traces are related, both are allowed to stutter with the empty event, and otherwise the events $\event$ and $\absevent$ must fulfill $\traceagree[]{\pi}{\event}{\absevent}$.
The trace agreement is the same for all properties $\pi$. 
With agreements, this section defines monitor satisfaction for traces and then it proves that monitor satisfaction implies property satisfaction.
To this end, monitor satisfaction is defined as follows.
\bul{A specification trace $\trace$ monitor-satisfies property $\pi$} iff there exists a (final) monitor state $\monitor$ and an abstract trace $\abstrace$ such that the \iul{specification trace $\trace$ agrees with abstract trace $\abstrace$} and \oul{the initial monitor\footnote{In this paper, for all monitors, the initial monitor state is denoted as $\emptyset$.} can step to the (final) monitor state $\monitor$ with abstract trace $\abstrace$}.

@@

% \begin{exampleenv}[A program not satisfying \gls*{tms}]\label{ex:strncpy:cpp11}
%   Consider the following \texttt{C++11} library that calls \texttt{strncpy} (\Cref{ex:strncpy}) and prints the result to the standard output stream.

%   \begin{lstlisting}[language=c++,basicstyle=\small\ttfamily,mathescape,commentstyle=\color{gray}]
% int hello() {
%   char* greet = new char[11] { "Hello CSF!" };
%   // greet address ${\color{gray}\loc_{x}}$
%   char* to = new char[11];
%   // to address ${\color{gray}\loc_{y}}$
%   strncpy(11, to, greetings);
%   delete to;
%   printf("%cSF\n", to[6]);
% }
%   \end{lstlisting}
%   Up to the body of \texttt{printf}, the program execution yields the specification trace $\ev{Alloc}\ \loc[_{x}]\ 12;\comp;\unlock\cdot\ev{Alloc}\ \loc[_{y}]\ 12;\comp;\unlock\cdot\ev{Use}\ \loc_{x}\ 0;\comp;\unlock\cdot\ev{Use}\ \loc_{x}\ 0;\comp;\unlock\cdot\ev{Use}\ \loc_{y}\ 0;\comp;\unlock\cdot\ev{Use}\ \loc_{x}\ 1;\comp;\unlock\cdot\ldots\cdot\ev{Use}\ \loc_{x}\ 12;\comp;\unlock\cdot\ev{Dealloc}\ \loc[_{y}];\comp;\unlock\cdot\ev{Use}\ \loc_{y}\ 6;\comp;\unlock$.
%   Relating this trace to abstract monitor events yields $\abstrace=\absev{Alloc}\ \loc[_{x}]\ \comp\cdot\absev{Alloc}\ \loc[_{y}]\ \comp\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Use}\ \loc[_{y}]\ \comp\cdot\ldots\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Dealloc}\ \loc[_{y}]\ \comp\cdot\absev{Use}\ \loc[_{y}]\ \comp$.
%   Remembering the definition of \texttt{strncpy} (\Cref{ex:strncpy}), observe that it does not deallocate its arguments.
%   Even though the trace contains an out-of-bounds access right before returning from \texttt{strncpy}, this is no concern for \gls*{tms}, since the location $\loc[_{x}]$ is still allocated.
%   However, having returned from \texttt{strncpy}, the \texttt{greet} function continues and deallocates $\loc[_{y}]$ whose subsequent use in the \texttt{printf} call is a use-after-free bug.

%   The fix would be to \lstinline[language=c++,basicstyle=\small\ttfamily]|delete greetings| instead of \texttt{to} and add a \lstinline[language=c++,basicstyle=\small\ttfamily]|delete to| after the \texttt{printf} call, which leads to the abstract monitor trace $\abstrace[']=\absev{Alloc}\ \loc[_{x}]\ \comp\cdot\absev{Alloc}\ \loc[_{y}]\ \comp\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Use}\ \loc[_{y}]\ \comp\cdot\ldots\cdot\absev{Use}\ \loc[_{x}]\ \comp\cdot\absev{Dealloc}\ \loc[_{x}]\ \comp\cdot\absev{Use}\ \loc[_{y}]\ \comp$.
%   It follows that $\montmssafe{\abstrace[']}$ and from \Thmref{lem:mon:tmssafe}, it follows that the program satisfies \Thmref{def:trace:tmsdef}, even though the program still violates \gls*{sms}.
% \end{exampleenv}

@@

% \begin{exampleenv}[Normal invocation of \texttt{strncpy}]\label{ex:strncpy:sms}
%   Consider the insecure \texttt{strncpy} function from \Cref{ex:strncpy} with a context \texttt{strncpy(2, x, y)}, where \texttt{x} and \texttt{y} are pointers to valid regions of memory with allocated space for exactly two cells and do not contain the null-terminating character \texttt{'\textbackslash 0'}.
%   For the sake of this example, the pointers have been allocated by the component and passed to the context.
%   The loop of \texttt{strncpy} will copy exactly two cells and then check the loop condition for the last time.
%   At that stage, the induction variable \texttt{i} is equal to \texttt{2} and, unfortunately, the order of checks is such that first the cell \texttt{x[i]} is read prior to bounds checking \texttt{i < n}.
%   Because of this, there is an out-of-bounds memory access right before exiting the function.
%   This is also visible on the trace, which can be sketched as $\ldots\cdot\ev{Alloc}\ \loc_{x}\ 2;\comp;\unlock\cdot\ldots\cdot\ev{Alloc}\ \loc_{y}\ 2;\comp;\unlock\cdot\ldots\cdot\ev{Use}\ \loc_{x}\ 0;\comp;\unlock\cdot\ev{Use}\ \loc_{y}\ 0;\comp;\unlock\cdot\ev{Use}\ \loc_{x}\ 1;\comp;\unlock\cdot\ev{Use}\ \loc_{y}\ 1;\comp;\unlock\cdot\ev{Use}\ \loc_{x}\ 2;\comp;\unlock\cdot\ldots$, where $\loc_{x}$ and $\loc_{y}$ are the memory addresses associated to \texttt{x} and \texttt{y}, respectively.
%   Omitting the events for all ,,$\ldots$'' for sake of brevity, the abstract monitor trace of this is $\absev{Alloc}\ \loc_{x}\ \comp\ 2\cdot\absev{Alloc}\ \loc_{y}\ \comp\ 2\cdot\absev{Use}\ \loc_{x}\ \comp\ 0\cdot\absev{Use}\ \loc_{y}\ \comp\ 0\cdot\absev{Use}\ \loc_{x}\ \comp\ 1\cdot\absev{Use}\ \loc_{y}\ \comp\ 1\cdot\absev{Use}\ \loc_{x}\ \comp\ 2$.

%   After the allocation events, the state of the monitor is $\left\{(\loc_{x};\comp;2),(\loc_{y};\comp;2)\right\}$.
%   All uses up to the last are accepted by the monitor, but the last event does not satisfy the premise $2<2$ in \Cref{tr:sms-use}.
%   Therefore, the whole program (\texttt{strncpy} linked with this kind of context) is not \gls*{sms}.
% \end{exampleenv}

@@

    \begin{tikzpicture}[scale=0.62, every node/.style={transform shape}]

@@

      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\large\ttfamily]

@@

      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\Large\ttfamily]

@@

\subsection{Secure Upper and Lower Composition}\label{sec:other-compos}
Besides sequential composition, there are two other compositions, namely an {\em upper}, i.e., a compiler that takes multiple inputs and yields one output, and a {\em lower} composition, i.e., a compiler that takes one input and yields multiple outputs.
\bul{Define the upper composition $\cc{\src{L}+\obj{L}}{\trg{L}}$} as follows:
Given a program \texttt{p}, its compiled counterpart is obtained by \rul{plugging \texttt{p} into $\cc{\src{L}}{\trg{L}}$ if $\texttt{p}\in\src{L}$} or by \oul{plugging \texttt{p} into $\cc{\obj{L}}{\trg{L}}$ if $\texttt{p}\in\obj{L}$}.
\begin{definition}[Upper Composition]
  $$\text{\bul{$\cc{\src{L}+\obj{L}}{\trg{L}}$}}\isdef
  % Given \texttt{p}, yield
  \lambda \texttt{p}\ldotp
  \left\{\mbox{\begin{tabular}{c}
    {\rul{if $\texttt{p}\in\src{L}$, then $\cc{\src{L}}{\trg{L}}(\texttt{p})$}} \\
    \mbox{\oul{if $\texttt{p}\in\obj{L}$, then $\cc{\obj{L}}{\trg{L}}(\texttt{p})$}} \\
  \end{tabular}}\right.$$
%
%
%  $\begin{cases}
%                                                                            \rul{\cc{\src{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\src{L}} \\
%                                                                            \cc{\obj{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\obj{L}
%                                                                          \end{cases}$
\end{definition}

@@

Dually, the {\em lower} composition is concerned about compilers that accept the same source but yield different target languages. %and the same theoretical results apply.
\bul{Define the lower composition $\cc{\src{L}}{\trg{L}+\obj{L}}$} as follows:
Given a program $\src{p}$, its compiled counterpart is obtained by \rul{plugging $\src{p}$ into $\cc{\src{L}}{\trg{L}}$} or by \oul{plugging \texttt{p} into $\cc{\src{L}}{\obj{L}}$}, respectively, \pul{based on the internal decision}.
\begin{definition}[Lower Composition]
  $$\text{\bul{$\cc{\src{L}}{\trg{L}+\obj{L}}$}}\isdef
  % Given $L$ and $\src{p}$, yield
  \lambda \src{p}, L\ldotp
  \left\{\mbox{\begin{tabular}{c}
    {\pul{if $L=\trg{L}$, then} \rul{$\cc{\src{L}}{\trg{L}}(\src{p})$}} \\
    \mbox{\pul{if $L=\obj{L}$, then} \oul{$\cc{\src{L}}{\obj{L}}(\src{p})$}} \\
  \end{tabular}}\right.$$
%
%
%  $\begin{cases}
%                                                                            \rul{\cc{\src{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\src{L}} \\
%                                                                            \cc{\obj{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\obj{L}
%                                                                          \end{cases}$
\end{definition}

@@

% This paper considers \gls*{cct} as an architecture-dependent property\MPin{can we justify this}, so
Writing code attaining \gls*{cct} should not be of the programmer's concerns~\cite{cauligi2019fact}.
Such consideration is also backed up by architecture providing a data (operand) independent timing mode, such as processors by Arm~\cite[p.~543]{arm-refman} and Intel~\cite[p.~80]{intel-refman}.
In spirit of this, all languages besides $\Lscct$ are trivially \gls*{scct}. 
$\Lscct$ (\Cref{subsec:lscct}) programs have access to a ``\gls*{cct}-mode'' and can change the leakage of emitted events according to the value of this mode (either $\obj{ON}$ or $\obj{OFF}$).

\subsection{Shared Language Definitions}\label{subsec:cs:defs}
\begin{gather*}
  \begin{aligned}
    \mi{(Expressions)}~&\expr\bnfdef\ x \mid \valueexpr \mid \lbinop{\expr[_{1}]}{\expr[_{2}]} \mid \lget{x}{\expr}\\
                       &\mid \llet{x}{\lnew{\expr[_{1}]}{\expr[_{2}]}}{\expr[_{3}]} \mid \ldelete{x}\\
                       &\mid \lset{x}{\expr[_{1}]}{\expr[_{2}]} \mid \lpair{\expr[_{1}]}{\expr[_{2}]} \mid \lproja{\expr} \mid \lprojb{\expr}\\
                       &\mid \llet{x}{\expr[_{1}]}{\expr[_{2}]} \mid \lreturn{\expr} \mid \lcall{g}{\expr}\\
                       &\mid \lifz{\expr}{\expr[_{1}]}{\expr[_{2}]} \mid \labort
  \end{aligned}
  \\
  \begin{aligned}
  \mi{(Functions)}~\function\bnfdef&\ \lfunction{\textit{foo}}{x}{e}
  \end{aligned}
  \\
  \begin{aligned}
  \mi{(Libraries)}~\library\bnfdef&\ \hole{\cdot} \mid \function,\library\hspace{0.5cm}
  \mi{(Programs)}~\prog{\library_{\ctx}}{\library_{\comp}}
  \end{aligned}
\end{gather*}
